Algebraic Principles for Rely-Guarantee Style Concurrency Verification Tools

### 2013-11-12 Paper/paper.tex
Almost finished paper

\DeclareFontFamily{U}{wncy}{}
\DeclareFontShape{U}{wncy}{m}{n}{<->wncyss10}{}
\DeclareSymbolFont{mcy}{U}{wncy}{m}{n}
\DeclareMathSymbol{\sh}{\mathord}{mcy}{"78} 
\DeclareMathOperator{\sha}{\sh}
\DeclareMathOperator{\tr}{Tr}
\DeclareMathOperator{\fin}{finite}
\DeclareMathOperator{\rely}{rely}
\DeclareMathOperator{\guar}{guar}
\DeclareMathOperator{\letters}{letters}
\DeclareMathOperator{\wlp}{wlp}
\DeclareMathOperator{\test}{test}
\DeclareMathOperator{\eval}{eval}
\DeclareMathOperator{\Id}{Id}
\DeclareMathOperator{\edn}{end}

@@

We aim to develop tools which support reasoning about concurrent
programs at varying levels of abstraction. At the highest level,
abstract algebra provides a framework for the prototyping of new
inference and refinement rules, whereas at the lowest level concrete
models of program stores allow for fine-grained reasoning about
program data flow. We have implemented this approach is the
Isabelle/HOL interactive theorem prover. Isabelle/HOL not only
provides us with a high degree of confidence in the correctness of our
theorems, but also supports the construction of custom proof tactics
and tools for program verification tasks. Our approach differs from
existing frameworks for rely-guarantee reasoning in
Isabelle~\cite{nieto}, both by making the link between concrete
programs and algebras explicit, but also by allowing arbitrary nested
parallelism.

For sequential programs the applicability of algebra, and Kleene
algebra in particular, has been known for over a decade. Kleene
algebra provides operations for non-deterministic choice, sequential
composition and finite iteration, in addition to skip and abort. With
appropriate extensions, Kleene algebras support Hoare-style program
verification, and allow the derivation of program equivalences and
refinement rules. Furthermore, Kleene algebra has been applied in many
applications, such as compiler
optimisation~\cite{kozen_certification_2000}, program construction,
transformation and termination analysis, and static analysis.

Algebra has also been applied in the study of separation logic, for
example, as the separation algebras in~\cite{ohearn}, or in algebraic
separation logic.

\fbox{Algebraic Separation Logic}

\fbox{Concurrent Kleene Algebra}

Rely-guarantee is a so called \emph{compositional} method, where the
verification of a large program is reduced the independent
verification of individual subprograms, with the interference from
other subprograms being abstractly captured by \emph{rely} and
\emph{guarantee} conditions. This form of compositionality is,
however, weaker than the strong notion of compositionality inherent in
algebra, where full congruence rules exist for every operator.

In this paper, we investigate algebraic principles for rely-guarantee
style reasoning. This has been explored previously, for example
in~\cite{cka}, but we extend this approach by considering a minimal
set of rules sufficient to derive the standard rely-guarantee
inference rules. Furthermore, we explore how the algebraic
compositionality of this approach can be broken in just the right way,
so as to capture the ideas of abstract interference conditions in
rely-guarantee reasoning.

The verification and construction of actual concurrent programs often
requires complex semantics involving synchronisation, termination
analysis and complicated data structures.

\fbox{We verify simple programs}

\fbox{Coherent framework for programs}

@@

\fbox{define weak trioid!}

@@

In the context of sequential programs, one typically thinks of $\cdot$
as sequential composition, $+$ as nondeterministic choice, $0$ as the
abortive action and $1$ as skip.  In this context it is essential that
multiplication is not commutative.  More formally, it is well known
that (regular) languages with language union as $+$ and language
product as $\cdot$, the empty language as $0$ and the empty word
language $\{\varepsilon\}$ as $1$ form dioids. Another model is formed
by binary relations with the union of relations as $+$, the product of
relations as $\cdot$, the empty relation as $0$ and the identity
relation as $1$. A model of commutative dioids is formed by sets of
(finite) multisets or Parikh vectors with multiset addition as
multiplication.

It is well known that commutative dioids can be used for modelling the
interaction between concurrent composition and nondeterministic
choice. The following definition serves as a basis for models of
concurrency in which sequential and concurrent composition
interact.\fbox{should we write $;$ instead of $\cdot$?}

A \emph{trioid} is a structure $(S,+,\cdot,||,0,1)$ such that
$(S,+,\cdot,0,1)$ is a dioid and $(S,\cdot,||,0,1)$ a commutative
dioid. In a trioid there is no interaction between the sequential
composition $\cdot$ and the parallel composition $||$. On the one
hand, Gischer has shown that trioids are sound and complete for the
equational theory of series-parallel pomset languages~\cite{Gischer},
which form a well studied model of true concurrency. On the other
hand, he has also obtained a completeness result with respect to a
notion of pomset subsumption for trioids with the additional
\emph{interchange axiom} $(w \| x) \cdot (y \| z) \le (w \cdot y) \|
(x \cdot z)$ and it is well known that this additional axiom also
holds for (regular) languages in which $||$ is interpreted as the
shuffle or interleaving operation~\cite{Gischer}.

Formally, the \emph{shuffle} $\sha$ of two finite words is defined
inductively as
\begin{equation*}
\epsilon \sha s = \{s\}, \qquad s \sha \epsilon = \{s\},\qquad
as \sha bt = a(s \sha bt) \cup b(as \sha t),
\end{equation*}
which is then lifted to the shuffle product of languages $X$ and $Y$
as
\begin{equation*}
X \sha Y = \{x \sha y:\; x \in X \land x \in Y\}.
\end{equation*}
\fbox{I would use $||$ here}

@@

It is well known that (regular) languages form Kleene algebras and
that (regular) sets of multisets form commutative Kleene algebras. In
fact, Kleene algebras are complete with respect to the equational
theory of regular languages as well as the equational theory of binary
relations with the reflexive transitive closure operation as the
star~\cite{Kozen}. Moreover, commutative Kleene algebras are
complete with respect to the equational theory of regular languages
over multisets~\cite{Conway}. It follows that equations in
(commutative) Kleene algebras are decidable.

A \emph{bi-Kleene algebra} is a structure
$(K,+,\cdot,||,0,1,^*,^{(*)})$ such that $(K,+,\cdot,0,1,^*)$ is a
Kleene algebra and $(K,+,||,0,1,^{(*)})$ a commutative Kleene
algebra. Bi-Kleene algebras are sound and complete with respect to the
equational theory of regular series-parallel pomset languages, and the
equational theory is again decidable~\cite{LaurenceStruth}. A
\emph{concurrent Kleene algebra} is a bi-Kleene algebra which
satisfies the interchange law~\cite{Hoareetal}. It can be shown that
shuffle languages and regular series-parallel pomset languages with a
suitable notion of pomset subsumption form concurrent Kleene algebras.

In some contexts, it is also useful to add a meet operation $\sqcap$,
such that $(D,+,\sqcap)$ is a lattice. \fbox{expand}

@@

It is well known that the inference rules for sequential Hoare logic
(except the assignment axiom) can be derived in expansions of Kleene
algebras. One approach is as follows~\cite{MoellerStruth}. Suppose a
suitable Boolean algebra $B$ of \emph{tests} has been embedded into a
Kleene algebra $K$ such that $0$ and $1$ are the minimal and maximal
element of $B$, $+$ corresponds to join and $\cdot$ to
meet. Complements $-$ are defined only on $B$. Suppose further that a
\emph{backward diamond operator} $\langle x|p$ has been defined for
each $x\in K$ and $p\in B$, which models the set of all states to
which each terminating execution of program $x$ may lead from states
$q$. Finally suppose that a \emph{forward box operator} $|x]p$ has
been defined which models the (largest) set of states from which every
terminating execution of $x$ must end up in states $q$ and that boxes
and diamonds are adjoints of the Galois connection
\begin{equation*}
  \langle x|p \le q \Leftrightarrow p \le |x]q,
\end{equation*}
for all $x\in K$ and $p,q\in B$. It is then evident from the above
explanations that validity of a Hoare triple $\vdash \{p\}x\{q\}$ can
be encoded as $\langle x|p \le q$ and the weakest liberal precodition
operator $\mathsf{wlp}(x,q)$ as $|x]p$. Hence the relationship between
the proof theory and the semantics of Hoare logic is captured by the
Galois connection $ \vdash\{p\}x\{q\} \Leftrightarrow p \le
\mathsf{wlp}(x,q)$. It has been shown that the relational semantics of
sequential while-programs can be encoded in these \emph{modal Kleene
  algebras} and that the inference rules of Hoare logic can be
derived~\cite{MoellerStruth}.

In the context of concurrency, this relational approach is no longer
appropriate. Here, the following approach by
Tarlecki~\cite{tarlecki_language_1985} can be used instead. One can
now encode validity of a Hoare triple as
\begin{equation*}
  \vdash \{x\}y\{z\} \Leftrightarrow x \cdot y \le z
\end{equation*}
for arbitrary elements of a Kleene algebra $K$. Nevertheless all the
rules of sequential Hoare logic except the assignment axiom can still
be derived in this setting~\cite{Hoareetal}.  Tarlecki's motivating
explanations carry over to the algebraic approach.

As an example we show the derivation of a generalised while
rule. Suppose $x\cdot t\cdot y\le x$. Then $x\cdot (t\cdot y)^\ast \le
x$ by the right induction axiom of Kleene algebra and therefore
$x\cdot (t\cdot y)^\ast t'\le x\cdot t'$ for arbitrary element $t'$ by
isotonicity of multiplication. This derives the while rule
\begin{equation*}
  \frac{\{x\cdot t\}y\{x\}}{\{x\}(t\cdot y)^\ast\{t'\cdot x\}}
\end{equation*}
for a generalised while loop $(t\cdot y)^\ast\cdot t'$, which
specialises to the conventional rule when $t$ and $t'$ are, in some
sense, complements.

The correspondence to a wlp-style semantics, as in model Kleene
algebra, now requires a generalisation of the Galois connection for
boxes and diamonds to multiplication and an upper adjoint in the form
of residuation.  This can be achieved in the context of \emph{action
  algebras}~\cite{pratt}, which expand Kleene algebras by operations
of left and right residuation defined by the Galois connections
\begin{equation*}
  xy\le z\Leftrightarrow x\le z\leftarrow y,\qquad xy\le z \Leftrightarrow y\le z\rightarrow x.
\end{equation*}
\fbox{please check that I didn't confuse these arrows} These
residuals, and now even the Kleene star can be axiomatised
equationally in action algebras. For a comprehensive list of the
properties of action algebras and their most important models
see~\cite{archive}, including the language and the relational
model. In analogy to the development in model Kleene algebra we can
now stipulate $ \wlp(x,y)=y\leftarrow x$ and obtain the Galois
connection
\begin{equation*}
  \vdash \{x\}y\{z\} \Leftrightarrow x\le \wlp(y,z)
\end{equation*}
with the characteristic properties $\vdash \{\wlp(y,z)\}y\{z\}$ and $x\le
\wlp(y,z)\Rightarrow \ \vdash \{x\}y\{z\}$. Moreover, if the action algebra is
also a quantale, and infinite sums exist, it follows that $
\wlp(y,z)=\sum\{x:\ \vdash \{x\}y\{z\}\}$.  It is obvious that this definition
makes sense in all models of action algebras and
quantales. Intuitively, suppose $p$ stands for the set of all
behaviours of a system, for instance the set of all execution traces,
that end in state $p$, and likewise for $q$. Then $\{p\}x\{q\}$ states
that all executions ending in $p$ can be extended by $x$ to executions
ending in $q$. $\wlp(x,q)$ is the most general behaviour, that is the
set of all executions $p$ after which all executions of $x$ must end
in $q$.

In the context of concurrent Kleene algebra, even without the parallel
star, additional inference rules can be derived, in particular 
\begin{equation*}
  \frac{\vdash\{x\}y\{z\}}{\vdash\{w||x\}y\{w||z\}}\qquad \frac{\vdash\{x_1\}y_1\{z_1\}\quad\vdash\{x_2\}y_2\{z_2\}}{\vdash\{x_1||x_2\}y_1||y_2\{z_1||z_2\}}
\end{equation*}
which correspond to the frame rule and the concurrency rule of
concurrent separation logic~\cite{} when $||$ is interpreted as
separating conjunction. In fact, the second inference rule is, in any
bi-Kleene algebra equivalent to the interchange axiom and the frame
rule is derivable from this axiom as well~\cite{Hoareetal}. The
precise connection between concurrent separation logic, a variant of
concurrent Kleene algebra and a resource transformer model has been
elaborated in~\cite{localitypaper}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{A  Rely-Guarantee Algebra}
\label{sec:RG}

In this section we show how bi-Kleene algebras can be expanded into a
simple algebra that supports the derivation of rely-guarantee-style
inference rules. This development does \emph{not} use the interchange
law for several reasons. First, this law fails for fair parallel
composition $x\parallel_f y$ in a model with possibly infinite, or
non-terminating programs. In this model, $x \cdot y \not\leq
x \parallel_f y$ in the case whenever $x$ is non-terminating. Second,
it is not needed for deriving the usual concurrency laws. Third,
whereas it may be assumed in the models outlined in Section *** it
fails for the Aczel trace model in Section ***. \fbox{is that true?}

\fbox{this needs to be reworked. two letters for a set are not good}

\fbox{ the set RG doesn't need to be named in the algebra I think.}

\fbox{we can just start with a bi-KA and say that elements are relies/guarantees if they satisfy properties}

\fbox{what is a weak trioid?}

\fbox{say that we don't consider the parallel star}

\fbox{should we call relies and guarantees more generically
  ``conditions'' can call them $c$?}

A rely-guarantee algebra is a structure
$(K,RG,+,\sqcap,\cdot,\|,^\star,0,1)$, where $(K,+,\sqcap)$ is a
distributive lattice, $(K,+,\cdot,\|,0,1)$ is a weak trioid and
$(K,+,\cdot,^\star,0,1)$ is a Kleene algebra. $RG$ is a distinguished
subset of relys and guarantees which satisfy the following axioms
\begin{align}
r\|r &\le r, \label{rg1}\\
r &\le r\|r', \label{rg2}\\
r\|xy &= (r\|x)(r\|y), \label{rg3}\\
r\|x^+ &\le (r\|x)^+ \label{rg4}.
\end{align}
By convention, we use $r$ and $g$ to refer to elements of $RG$, and
$x,y,z$ for arbitrary elements of $K$.

@@

Some elementary consequences of these rules are as follows
\begin{align*}
1 &\le r,\\
r^\star = rr &= r = r\|r,\\
r\|x^+ &= (r\|x)^+.
\end{align*}

%% Could use better explanations.

Axioms (\ref{rg1}), (\ref{rg2}) and (\ref{rg3}) are independent. To
show this, we use Isabelle's
\emph{Nitpick}~\cite{blanchette_nitpick:_2010} counterexample generator
to construct models which violate each of these axioms, yet satisfy
all others. The models thus constructed for (\ref{rg1}) and
(\ref{rg3}) are shown in Figures \ref{fig:rg1} and \ref{fig:rg3}
respectively. The model constructed for (\ref{rg2}) is just the two
element rely-guarantee algebra with $0$ and $1$.

\begin{figure}[H]
\centering
\begin{minipage}{0.24\textwidth}
\begin{tikzpicture}[x=1.5cm,y=1.5cm,auto]
  \node (center) {};
  \node [elem] (r1) at (90:1) {$r_1$};
  \node [elem] (one) at (200:1) {$1$};
  \node [elem] (r2) at (20:1) {$r_2$};
  \node [elem] (zero) at (-90:1) {$0$};

  \path [line] (r1) -- (r2);
  \path [line] (r1) -- (one);
  \path [line] (r1) -- (zero);
  \path [line] (r2) -- (one);
  \path [line] (r2) -- (zero);
  \path [line] (one) -- (zero);
\end{tikzpicture}
\end{minipage}
\begin{minipage}{0.24\textwidth}
\begin{align*}
r_1 \| r_1 &= r_1\\
r_1 \| r_2 &= r_1\\
r_2 \| r_1 &= r_1\\
r_2 \| r_2 &= r_1
\end{align*}
\end{minipage}
\begin{minipage}{0.24\textwidth}
\begin{align*}
r_1r_1 &= r_1\\
r_1r_2 &= r_1\\
r_2r_1 &= r_1\\
r_2r_2 &= r_2
\end{align*}
\end{minipage}
\begin{minipage}{0.24\textwidth}
\begin{align*}
r_1^\star &= r_1\\
r_2^\star &= r_2
\end{align*}
\end{minipage}
\caption{4 element counterexample for $r \in RG \implies r\|r \le r$}
\label{fig:rg1}
\end{figure}

Axiom (\ref{rg4}) can be derived from (\ref{rg3}) in all finite
models. This is because any finite $K$ is complete, and in a complete
setting fixpoint laws can be used to show $r\|x^+ \le (r\|x)^+$. Thus
it is impossible to construct a finite model demonstrating that
(\ref{rg4}) is independent from (\ref{rg1}) -- (\ref{rg3}).

\begin{figure}[t]
\centering
\begin{minipage}{0.24\textwidth}
\begin{tikzpicture}[x=1.5cm,y=1.5cm,auto]
  \node (center) {};
  \node [elem] (r1) at (90:1) {$r_1$};
  \node [elem] (one) at (0:1) {$1$};
  \node [elem] (zero) at (-90:1) {$0$};

  \path [line] (r1) -- (one);
  \path [line] (r1) -- (zero);
  \path [line] (r1) -- (zero);
  \path [line] (one) -- (zero);
\end{tikzpicture}
\end{minipage}
\begin{minipage}{0.24\textwidth}
\begin{align*}
r_1 \| r_1 &= r_1\\
r_1 r_1 &= r_1\\
r_1^\star &= r_1
\end{align*}
\end{minipage}
\caption{3 element counterexample for $r \in RG \implies r\|xy = (r\|x)(r\|y)$}
\label{fig:rg3}
\end{figure}

@@

By using these residuals quintuples can be encoded in the following way,
which is equivalent to the encoding in Equation (\ref{quin}).
\begin{align}
r, g \vdash \{p\} c \{q\} \iff c \le r/(p \rightarrow q) \sqcap g \label{refine}.
\end{align}
This encoding allows us to think in terms of program refinement, as
$r/(p \rightarrow q) \sqcap g$ defines the weakest program that when
placed in parallel with interference from $r$, and guaranteeing
interference at most $g$, goes from $p$ to $q$---a generic
specification for a concurrent program.

@@

% Or should this be weak?

>>>>

\DeclareFontFamily{U}{wncy}{}
\DeclareFontShape{U}{wncy}{m}{n}{<->wncyss10}{}
\DeclareSymbolFont{mcy}{U}{wncy}{m}{n}
\DeclareMathSymbol{\sh}{\mathord}{mcy}{"78}
\DeclareMathOperator{\sha}{\sh}
\DeclareMathOperator{\tr}{Tr}
\DeclareMathOperator{\fin}{finite}
\DeclareMathOperator{\rely}{rely}
\DeclareMathOperator{\guar}{guar}
\DeclareMathOperator{\letters}{letters}

\newcommand{\wlp}{\mathsf{wlp}}

\newcommand{\test}{\mathsf{test}}
\newcommand{\eval}{\mathsf{eval}}
\newcommand{\Id}{\mathsf{Id}}
\newcommand{\edn}{\mathsf{end}}
\newcommand{\unchanged}{\mathsf{unchanged}}
\newcommand{\preserves}{\mathsf{preserves}}

@@

We aim to develop lightweight formal methods which support
rely-guarantee style reasoning about concurrent programs at varying
levels of abstraction. At the most abstract level, we use algebras as
a framework for reasoning about the control flow of programs as well
as for deriving inference rules and refinement laws. At the lowest
level, concrete models of program stores support fine-grained
reasoning about program data flow and inteference. We have implemented
this approach is the Isabelle/HOL interactive theorem proving
environment. Isabelle not only provides us with a high degree of
confidence in the correctness of our development and tool prototype,
it also supports the construction of custom proof tactics and
procedures for program verification and refinement tasks. Our approach
differs from existing frameworks for rely-guarantee reasoning in
Isabelle~\cite{nieto_rely-guarantee_2003}, both by making the link
between concrete programs and algebras explicit, and by allowing
arbitrary nested parallelism.

For sequential programs, the applicability of algebra, and Kleene
algebra in particular, has been known for over a decade. Kleene
algebra provides operations for non-deterministic choice, sequential
composition and finite iteration, in addition to skip and abort. With
appropriate extensions, Kleene algebras support Hoare-style program
verification, and allow the derivation of program equivalences and
refinement rules~\cite{kozen_kleene_1997,hoare_concurrent_2011}. Kleene algebras have
been applied in applications including compiler
optimisation~\cite{kozen_certification_2000}, program
construction~\cite{berghammer_automated_2010}, transformation and
termination analysis~\cite{desharnais_algebraic_2011}, and static
analysis~\cite{fernandes_describing_2007}. Formalisations and tools are
available in interactive theorem provers such as Coq~\cite{pous_kleene_2013} and
Isabelle~\cite{armstrong_kleene_2013,armstrong_program_2013,armstrong_algebras_2014}.

Algebra also plays a role in the study of separation logic, for
example, as the separation algebras in~\cite{calcagno_local_2007}, and in algebraic
separation logic~\cite{dang_algebraic_2011}. More recently,
concurrent Kleene algebras have given an algebraic account of some
aspects of concurrent separation logic~\cite{hoare_concurrent_2011,hoare_locality_2011} and
the first steps towards an algebraic description of rely-guarantee
based reasoning have been undertaken~\cite{hoare_concurrent_2011}.

One main characteristic of the rely-guarantee method is
compositionality. The verification of large concurrent programs is
reduced to the independent verification of individual subprograms,
where the effect of interferences from other subprograms are captured
by \emph{rely} and \emph{guarantee} conditions, which constrain the
behaviour of subprograms. This form of compositionality is, however,
weaker than the strong notion of compositionality inherent in algebra,
where full congruence rules exist for every operator and refinement is
usually based on monotonicity.

In this paper, we investigate algebraic principles for rely-guarantee
style reasoning. First step have already been taken, for example,
in~\cite{hoare_concurrent_2011}, but we extend this approach by extracting a minimal set
of axioms for relies and guarantees sufficient to derive the standard
rely-guarantee inference rules. Furthermore, we explore how the
algebraic compositionality of this approach can be broken in just the
right way, so as to capture the ideas of abstract interference
conditions in rely-guarantee reasoning.

The verification and construction of actual concurrent programs often
requires complex semantics involving synchronisation, termination
analysis and intricate data structures. For simplicity, in this paper
we eschew such models in favor of relatively simple language and trace
based semantics. Despite the simplicity of this model, we demonstrate
and evaluate our prototypical verification tool implemented in
Isabelle by verifiying simple two examples from the literature. Beyond
that our approach provides a coherent framework from which more
complex and detailed models can be integrated in the future.

@@

@@

In the context of sequential programs, one typically thinks of $\cdot$
as sequential composition, $+$ as nondeterministic choice, $0$ as the
abortive action and $1$ as skip. In this context it is essential that
multiplication is not commutative. Often we use $;$ in place of
$\cdot$ when discussing programs. More formally, it is well known that
(regular) languages with language union as $+$ and language product as
$\cdot$, the empty language as $0$ and the empty word language
$\{\varepsilon\}$ as $1$ form dioids. Another model is formed by
binary relations with the union of relations as $+$, the product of
relations as $\cdot$, the empty relation as $0$ and the identity
relation as $1$. A model of commutative dioids is formed by sets of
(finite) multisets or Parikh vectors with multiset addition as
multiplication.

It is well known that commutative dioids can be used for modelling the
interaction between concurrent composition and nondeterministic
choice. The following definition serves as a basis for models of
concurrency in which sequential and concurrent composition
interact.

A \emph{trioid} is a structure $(S,+,\cdot,||,0,1)$ such that
$(S,+,\cdot,0,1)$ is a dioid and $(S,\cdot,||,0,1)$ a commutative
dioid. In a trioid there is no interaction between the sequential
composition $\cdot$ and the parallel composition $||$. On the one
hand, Gischer has shown that trioids are sound and complete for the
equational theory of series-parallel pomset languages~\cite{gischer_equational_1988},
which form a well studied model of true concurrency. On the other
hand, he has also obtained a completeness result with respect to a
notion of pomset subsumption for trioids with the additional
\emph{interchange axiom} $(w \| x) \cdot (y \| z) \le (w \cdot y) \|
(x \cdot z)$ and it is well known that this additional axiom also
holds for (regular) languages in which $||$ is interpreted as the
shuffle or interleaving operation~\cite{gischer_shuffle_1981}.

Formally, the \emph{shuffle} $\|$ of two finite words is defined
inductively as
\begin{equation*}
\epsilon \| s = \{s\}, \qquad s \| \epsilon = \{s\},\qquad
as \| bt = a(s \| bt) \cup b(as \| t),
\end{equation*}
which is then lifted to the shuffle product of languages $X$ and $Y$
as
\begin{equation*}
X \| Y = \{x \| y:\; x \in X \land x \in Y\}.
\end{equation*}

@@

It is well known that (regular) languages form Kleene algebras and
that (regular) sets of multisets form commutative Kleene algebras. In
fact, Kleene algebras are complete with respect to the equational
theory of regular languages as well as the equational theory of binary
relations with the reflexive transitive closure operation as the
star~\cite{kozen_completeness_1994}. Moreover, commutative Kleene
algebras are complete with respect to the equational theory of regular
languages over multisets~\cite{conway_regular_1971}. It follows that
equations in (commutative) Kleene algebras are decidable.

A \emph{bi-Kleene algebra} is a structure
$(K,+,\cdot,||,0,1,\hphantom{}^*,\hphantom{}^{(*)})$ where
$(K,+,\cdot,0,1,\hphantom{}^*)$ is a Kleene algebra and
$(K,+,||,0,1,\hphantom{}^{(*)})$ is a commutative Kleene
algebra. Bi-Kleene algebras are sound and complete with respect to the
equational theory of regular series-parallel pomset languages, and the
equational theory is again decidable~\cite{LaurenceStruth}. A
\emph{concurrent Kleene algebra} is a bi-Kleene algebra which
satisfies the interchange law~\cite{hoare_concurrent_2011}. It can be shown that
shuffle languages and regular series-parallel pomset languages with a
suitable notion of pomset subsumption form concurrent Kleene algebras.

In some contexts, it is also useful to add a meet operation $\sqcap$,
such that $(D,+,\sqcap)$ is a distributive lattice. This is
particularly needed in the context of refinement, where we typically
want to represent specifications as well as programs.

@@

It is well known that the inference rules for sequential Hoare logic
(except the assignment axiom) can be derived in expansions of Kleene
algebras. One approach is as follows~\cite{moller_algebras_2006}. Suppose a
suitable Boolean algebra $B$ of \emph{tests} has been embedded into a
Kleene algebra $K$ such that $0$ and $1$ are the minimal and maximal
element of $B$, $+$ corresponds to join and $\cdot$ to
meet. Complements $-$ are defined only on $B$. Suppose further that a
\emph{backward diamond operator} $\langle x|p$ has been defined for
each $x\in K$ and $p\in B$, which models the set of all states to
which each terminating execution of program $x$ may lead from states
$q$. Finally suppose that a \emph{forward box operator} $|x]p$ has
been defined which models the (largest) set of states from which every
terminating execution of $x$ must end up in states $q$ and that boxes
and diamonds are adjoints of the Galois connection
\begin{equation*}
  \langle x|p \le q \Leftrightarrow p \le |x]q,
\end{equation*}
for all $x\in K$ and $p,q\in B$. It is then evident from the above
explanations that validity of a Hoare triple $\vdash \{p\}x\{q\}$ can
be encoded as $\langle x|p \le q$ and the weakest liberal precodition
operator $\mathsf{wlp}(x,q)$ as $|x]p$. Hence the relationship between
the proof theory and the semantics of Hoare logic is captured by the
Galois connection $ \vdash\{p\}x\{q\} \Leftrightarrow p \le
\mathsf{wlp}(x,q)$. It has been shown that the relational semantics of
sequential while-programs can be encoded in these \emph{modal Kleene
  algebras} and that the inference rules of Hoare logic can be
derived~\cite{moller_algebras_2006}.

In the context of concurrency, this relational approach is no longer
appropriate. Here, the following approach by
Tarlecki~\cite{tarlecki_language_1985} can be used instead. One can
now encode validity of a Hoare triple as
\begin{equation*}
  \vdash \{x\}y\{z\} \Leftrightarrow x \cdot y \le z
\end{equation*}
for arbitrary elements of a Kleene algebra $K$. Nevertheless all the
rules of sequential Hoare logic except the assignment axiom can still
be derived in this setting~\cite{hoare_concurrent_2011}. Tarlecki's motivating
explanations carry over to the algebraic approach.

As an example we show the derivation of a generalised while
rule. Suppose $x\cdot t\cdot y\le x$. Then $x\cdot (t\cdot y)^\ast \le
x$ by the right induction axiom of Kleene algebra and therefore
$x\cdot (t\cdot y)^\ast t'\le x\cdot t'$ for arbitrary element $t'$ by
isotonicity of multiplication. This derives the while rule
\begin{equation*}
  \frac{\vdash \{x\cdot t\}y\{x\}}{\vdash \{x\}(t\cdot y)^\ast\{t'\cdot x\}}
\end{equation*}
for a generalised while loop $(t\cdot y)^\ast\cdot t'$, which
specialises to the conventional rule when $t$ and $t'$ are, in some
sense, complements.

The correspondence to a wlp-style semantics, as in model Kleene
algebra, now requires a generalisation of the Galois connection for
boxes and diamonds to multiplication and an upper adjoint in the form
of residuation.  This can be achieved in the context of \emph{action
  algebras}~\cite{pratt_action_1990}, which expand Kleene algebras by operations
of left and right residuation defined by the Galois connections
\begin{equation*}
  x\cdot y\le z\Leftrightarrow x\le z\leftarrow y,\qquad x\cdot y\le z \Leftrightarrow y\le x\rightarrow z.
\end{equation*}
These residuals, and now even the Kleene star can be axiomatised
equationally in action algebras. For a comprehensive list of the
properties of action algebras and their most important models
see~\cite{armstrong_kleene_2013}, including the language and the relational
model. In analogy to the development in model Kleene algebra we can
now stipulate $ \wlp(x,y)=y\leftarrow x$ and obtain the Galois
connection
\begin{equation*}
  \vdash \{x\}y\{z\} \Leftrightarrow x\le \wlp(y,z)
\end{equation*}
with the characteristic properties $\vdash \{\wlp(y,z)\}y\{z\}$ and $x\le
\wlp(y,z)\Rightarrow \ \vdash \{x\}y\{z\}$. Moreover, if the action algebra is
also a quantale, and infinite sums exist, it follows that $
\wlp(y,z)=\sum\{x:\ \vdash \{x\}y\{z\}\}$.  It is obvious that this definition
makes sense in all models of action algebras and
quantales. Intuitively, suppose $p$ stands for the set of all
behaviours of a system, for instance the set of all execution traces,
that end in state $p$, and likewise for $q$. Then $\{p\}x\{q\}$ states
that all executions ending in $p$ can be extended by $x$ to executions
ending in $q$. $\wlp(x,q)$ is the most general behaviour, that is the
set of all executions $p$ after which all executions of $x$ must end
in $q$.

In the context of concurrent Kleene algebra, even without the parallel
star, additional inference rules can be derived, in particular
\begin{equation*}
  \frac{\vdash\{x\}y\{z\}}{\vdash\{w||x\}y\{w||z\}}\qquad \frac{\vdash\{x_1\}y_1\{z_1\}\quad\vdash\{x_2\}y_2\{z_2\}}{\vdash\{x_1||x_2\}y_1||y_2\{z_1||z_2\}}
\end{equation*}
which correspond to the frame rule and the concurrency rule of
concurrent separation logic~\cite{} when $||$ is interpreted as
separating conjunction. In fact, the second inference rule is, in any
bi-Kleene algebra equivalent to the interchange axiom and the frame
rule is derivable from this axiom as well~\cite{hoare_concurrent_2011}. The
precise connection between concurrent separation logic, a variant of
concurrent Kleene algebra and a resource transformer model has been
elaborated in~\cite{hoare_locality_2011}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{A  Rely-Guarantee Algebra}
\label{sec:RG}

In this section we show how bi-Kleene algebras can be expanded into a
simple algebra that supports the derivation of rely-guarantee-style
inference rules. This development does \emph{not} use the interchange
law for several reasons. First, this law fails for fair parallel
composition $x\parallel_f y$ in a model with possibly infinite, or
non-terminating programs. In this model, $x \cdot y \not\leq x
\parallel_f y$ in the case whenever $x$ is non-terminating. Secondly,
it is not needed for deriving the usual laws or rely-guarantee.

A rely-guarantee algebra is a structure
$(K,C,+,\sqcap,\cdot,\|,^\star,0,1)$, where $(K,+,\sqcap)$ is a
distributive lattice, $(K,+,\cdot,\|,0,1)$ is a trioid and
$(K,+,\cdot,^\star,0,1)$ is a Kleene algebra. Note that we do not
consider the parallel star. $C$ is a distinguished subset of rely and
guarantee conditions or \emph{interference constraints} which satisfy
the following axioms
\begin{align}
r\|r &\le r, \label{rg1}\\
r &\le r\|r', \label{rg2}\\
r\|xy &= (r\|x)(r\|y), \label{rg3}\\
r\|x^+ &\le (r\|x)^+ \label{rg4}.
\end{align}
By convention, we use $r$ and $g$ to refer to elements of $C$,
depending on whether they are used as relies or guarantees, and
$x,y,z$ for arbitrary elements of $K$. The operators $\|$ and $\sqcap$
must be closed with respect to $C$.

@@

Some elementary consequences of these rules are
\begin{equation*}
1 \le r,\qquad
r^\star = rr = r = r\|r,\qquad
r\|x^+ = (r\|x)^+.
\end{equation*}

Axioms (\ref{rg1}), (\ref{rg2}) and (\ref{rg3}) are independent. To
show this, we use Isabelle's
\emph{Nitpick}~\cite{blanchette_nitpick:_2010} counterexample
generator to construct models which violate each particular axiom
while satisfying all others.

Axiom (\ref{rg4}) can be derived from (\ref{rg3}) by a simple
inductive argument in quantales where the plus operation can be
defined as a sum of powers $x^+=\sum_{i\ge 1} x^i$ where $x^1=x$
and $x^{i+1}=x\cdot x^i$. In first-order Kleene algebras, however, it
is independent, but it is impossible to find a counterexample with Nitpick because all such counterexamples are finite, but all finite Kleene algebras are a forteriori quantales.

@@

By using these residuals quintuples can be encoded in the following way,
which is equivalent to the encoding in Equation (\ref{quin}).
\begin{align}
r, g \vdash \{p\} c \{q\} \iff c \le r/(p \rightarrow q) \sqcap g \label{refine}.
\end{align}
This encoding allows us to think in terms of program refinement~\cite{hayes_refining_2013}, as
$r/(p \rightarrow q) \sqcap g$ defines the weakest program that when
placed in parallel with interference from $r$, and guaranteeing
interference at most $g$, goes from $p$ to $q$---a generic
specification for a concurrent program.

@@
